---
title: "The R functions, by example"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

zufast is a C library first: its R functions exist to test, benchmark and
demonstrate the headers, and they are a convenient way to see what the C layer
does. This article runs every exported function. All of them are vectorised,
keep `NA` as `NA`, and turn invalid input into `NA` (or `NULL` in a list)
without a warning. Arguments of the wrong type raise a condition of class
`zufast_invalid_argument`, which inherits from `zufast_error`.

```{r setup}
library(zufast)
```

## Package information

`fast_info()` reports, from compiled code, the version of the headers zufast's
own shared object was built with, the vendored library versions and the
compiler.

```{r}
str(fast_info())
```

## Numbers

### `fast_parse_double()`

Decimal and scientific notation, `Inf`, `Infinity` and `NaN` in any case, with
a leading `-` or `+`, after trimming ASCII whitespace. Anything else, including
trailing characters and hexadecimal, is `NA`.

```{r}
fast_parse_double(c("1.5", " -2e-3 ", "+inf", "nan", ".5", "1.", "1e400", "1e-400"))
fast_parse_double(c("0x10", "1,5", "12abc", "", NA))
```

The result is correctly rounded: the nearest double, ties to even. R's
`as.numeric()` is not, and the two can differ in the last bit:

```{r}
x <- "0.799012"
fast_parse_double(x) == as.numeric(x)
sprintf("%a", c(fast_parse_double(x), as.numeric(x)))
```

### `fast_parse_integer()`

Decimal integers only, in R's integer range. Unlike `as.integer()`, nothing
goes through a double, so a fraction, an exponent or an overflow is `NA`
rather than truncated.

```{r}
fast_parse_integer(c("42", " +7 ", "-2147483647", "2147483648", "1.0", "1e3", "0x1A"))
as.integer(c("1.9", "1e3"))   # base R, for comparison
```

### `fast_format_double()`

The fewest significant digits that read back to exactly the same double, in
ECMAScript `Number.prototype.toString()` notation.

```{r}
x <- c(1, 0.1, 0.1 + 0.2, 1 / 3, 1e21, 1e-7, 123456.789, -0, NA, NaN, Inf)
fast_format_double(x)
as.character(x)               # base R: fifteen significant digits
identical(fast_parse_double(fast_format_double(x)), x)
```

`scientific = TRUE` always writes `d.ddde+x`; `trailing_zero = TRUE` keeps
integral values recognisably floating point.

```{r}
fast_format_double(c(1, 123.456, 1e-3), scientific = TRUE)
fast_format_double(c(1, 2.5, 1e20), trailing_zero = TRUE)
```

## Dates and timestamps

### `fast_parse_date()`

`YYYY-MM-DD` with full validation in the proleptic Gregorian calendar.

```{r}
fast_parse_date(c("2024-02-29", "2023-02-29", "1900-02-29", "2000-02-29",
                  "2024-13-01", "2024-1-5", "2024-01-01T00:00", NA))
```

### `fast_parse_datetime()`

RFC 3339 and the common ISO 8601 forms: `T`, `t` or a space between date and
time, optional seconds, 1 to 9 fraction digits, and `Z` or a numeric offset.
The result is a `POSIXct` in UTC; an offset is applied, and a timestamp
without one is taken to be UTC. A bare date is midnight.

```{r}
x <- c("2024-03-01T12:30:00Z",
       "2024-03-01 12:30:00.25+01:00",
       "2024-03-01t12:30-0530",
       "2016-12-31T23:59:60Z",       # leap second: the next minute
       "2024-03-01",
       "2024-03-01T24:00:00Z",       # rejected
       "2024-03-01T12:30:00,5Z")     # comma fractions are rejected
fast_parse_datetime(x)
```

### `fast_datetime_fields()`

The parsed fields themselves, for a caller that wants to apply a time zone of
its own or tell "no offset" apart from "UTC".

```{r}
fast_datetime_fields(c("2024-03-01T12:30:00.123456789-05:00",
                       "2024-03-01T12:30:00",
                       "2024-03-01",
                       "not a date"))
```

### `fast_format_datetime()`

RFC 3339 in UTC for `POSIXct`, `YYYY-MM-DD` for `Date`.

```{r}
t <- as.POSIXct(c("2024-03-01 12:30:00", "2024-03-01 12:30:00.25"), tz = "UTC")
fast_format_datetime(t)
fast_format_datetime(t, digits = 0)
fast_format_datetime(t, digits = 6)
fast_format_datetime(as.Date(c("2024-03-01", "0001-01-01", NA)))
```

Formatting and parsing round-trip:

```{r}
t <- as.POSIXct("1969-07-20 20:17:40", tz = "UTC")
identical(fast_parse_datetime(fast_format_datetime(t)), t)
```

## Bytes and text

### `fast_hex_encode()` and `fast_hex_decode()`

A raw vector encodes to one string; a character vector encodes each element's
bytes. Decoding accepts either case and returns a list of raw vectors, `NULL`
where the input is `NA` or not hex.

```{r}
fast_hex_encode(as.raw(c(0xde, 0xad, 0xbe, 0xef)))
fast_hex_encode(c("zu", "fast", NA), upper = TRUE)
fast_hex_decode(c("DEADbeef", "abc", "zz", NA))
```

### `fast_base64_encode()` and `fast_base64_decode()`

Standard or URL-safe alphabet, with or without padding.

```{r}
fast_base64_encode(c("f", "fo", "foo", "foob"))
fast_base64_encode(as.raw(c(0xfb, 0xff, 0xbf)), url = TRUE)
fast_base64_encode("hello", url = TRUE, pad = FALSE)
```

Decoding is strict, so that every accepted input has exactly one encoding:
whitespace, the other alphabet, partial padding and non-zero unused bits are
all rejected. Padding itself is optional.

```{r}
decoded <- fast_base64_decode(c("aGVsbG8=", "aGVsbG8", "aGVsbG9=", "aGVs bG8=", NA))
lapply(decoded, function(r) if (is.null(r)) NULL else rawToChar(r))
fast_base64_decode("-_-_", url = TRUE)
```

### `fast_hash()`

XXH3, 64 or 128 bits, as lower-case hex in xxHash's canonical form. The values
match the reference implementation. **XXH3 is not a cryptographic hash**: use
it for caches, deduplication and change detection, never against an
adversary.

```{r}
fast_hash(charToRaw("hello"))
fast_hash(c("hello", "world", NA))
fast_hash("hello", bits = 128)
fast_hash("hello", seed = 42)
```

A string and its bytes hash alike:

```{r}
fast_hash("hello") == fast_hash(charToRaw("hello"))
```

### `fast_utf8_valid()`

Well-formed UTF-8: no overlong forms, no surrogates, nothing above U+10FFFF.
The declared encoding is ignored; only the bytes are examined.

```{r}
cafe <- rawToChar(as.raw(c(0x63, 0x61, 0x66, 0xc3, 0xa9)))   # "cafe" with an acute e
latin1 <- rawToChar(as.raw(c(0x63, 0x61, 0x66, 0xe9)))       # the same in Latin-1
fast_utf8_valid(c("plain", cafe, latin1, NA))
fast_utf8_valid(as.raw(c(0xc0, 0xaf)))         # overlong "/"
fast_utf8_valid(as.raw(c(0xed, 0xa0, 0x80)))   # a surrogate
```

## Errors

Argument problems are classed conditions, so they can be caught precisely:

```{r}
tryCatch(fast_parse_double(1.5), zufast_invalid_argument = function(e) conditionMessage(e))
tryCatch(fast_hash("x", bits = 32), zufast_error = function(e) class(e))
```
